Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-11-14 paper/main.tex
acks
more writing

\newcommand{\supplement}{~(\href{http://doi.org/10.5281/zenodo.4167148}{doi:10.5281/zenodo.4167148})}

@@

\subsection{Distributions over permutations}
Distributions over permutations~\cite{critchlow91} are functions that assign a probability value  $\Prob(\sigma)\in[0,1]$ to each permutation $\sigma \in S_n$. One of the most popular distributions is the Mallows Model (MM), which is considered as an analogous to the Gaussian distribution for permutations. The MM defines the probability of each permutation $\sigma\in S_n$ as follows:
%
\begin{equation}\label{eq:MM}
\Prob(\sigma \mid \sigma_0, \theta )= \frac{\exp(-\theta d(\sigma, \sigma_0))}{\psi} \sim MM(\sigma_0, \theta)
\end{equation}
%
with two parameters, $\theta$ and $\sigma_0$. Permutation $\sigma_0\in S_n$ is a reference permutation and has the largest probability value, i.e., the mode of the distribution. The probability of every permutation $\sigma\in S_n$ decays exponentially as its distance $d(\sigma,\sigma_0)$ increases, and the dispersion parameter $\theta$  controls this decay.  The normalization constant $\psi$ can be easily computed for the Kendall's $\tau$ distance (Eq.~\eqref{eq:kendall_decomp}) as well as for the Hamming, Cayley and Ulam distances~\cite{IruCalLoz2016permallows}.

@@

The exact MLE for $\sigma_0$ is given by the well-known Kemeny ranking~\cite{DwoKumNao2001rank}. Unfortunately,  obtaining such ranking is NP-hard~\cite{DwoKumNao2001rank}. An alternative to the Kemeny ranking is the Borda ranking~\cite{AliMei2011kemeny}, which has several advantages: (i) it requires polynomial computational time; (ii) it  guaranties  high quality parameters for a sample distributed according to the MM~\cite{CarProSha2013votes}; and (iii) and also when no distribution is assumed, Borda is an approximation to Kemeny~\cite{CopFleRur2010ordering}.

@@

\subsubsection{Datasets}
%
For the LOP, we generated synthetic instances as follows. First, we create a sample $S$ of 200 permutations of size $n=20$ from a MM with $\sigma_0$ equal to the identity permutation and $\theta=-\log(\textit{Uniformity})$, where  $\textit{Uniformity}\in\{0.5,$ $0.7,$ $0.9\}$, i.e., from peaked distributions to almost uniform distributions. Then, we generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbf{I}\{\sigma(i)<\sigma(j)\}$, where $\mathbf{I}\colon S \times S \to \{0,1\}$ is the indicator function. In principle, these should be best-case instances for UMM since their fitness landscape closely matches the Mallows model.\MANUEL{Ekhine: Could you revise this text? I wanted to explain why we use synthetic datasets.}

@@

The first part of the experimental analysis concerns the {analyses} of the learning rate, which corresponds to parameter $\rho$ of uBorda. As stated in Section~\ref{sec:sampling}, $\rho$ is automatically set at each iteration in such a way that the $100r_1\%$ of the samples have the $100r_2\%$ of the weight. In this regard, we tried a few different values of parameters $r_1$ and $r_2$ in all the instances. Only values of $r_1 < r_2$ make sense and $r_1 = r_2 = 0.5$ would produce equal weights in the learning step. Thus, we tried values $r_1 = \{0.1, 0.2, \dotsc, 0.5\}$ and $r_2 = \{0.6,\dotsc, 0.9, 0.99\}$. Figure~\ref{fig:heatmaps} shows the best solution found by each independent run (averaged over 10 runs) for instances: \texttt{N-t1d100.01} and \texttt{N-t1d150.01} for the LOP, \texttt{reC05} for the PFSP, \texttt{kra32} for the QAP. We terminated runs that took longer than 6 hours and those combinations are shown as white.  Results for all the datasets can be found in the supplemental material\supplement.

@@

The situation is very similar for the QAP, where both algorithms show the same
behaviour as in the PFSP for all instances tested, except for
\texttt{kra32}. As shown in Fig~\ref{fig:kra32}, for \texttt{kra32} both
algorithms are able to improve over the number of evaluations, although not
with ease. As in the synthetic LOP instances, we again observe that CEGO does
improve faster than UMM initially, yet CEGO appears to
get stuck around evaluation 250,  and UMM overtakes it around evaluation 300. We believe the reason is
the same as before, that is, the CEGO model is not able to adequately reflect the
ruggedness of the actual landscape and, hence, the best solutions found by the
GA searching the model are not improved solutions for the actual problem. Interestingly, \texttt{kra32} is rectangular
whereas \texttt{tho32} contains rectangular distances and \texttt{nug12} and \texttt{nug30} contain Manhattan distances of rectangular grids.\MANUEL{so they all have the same distance type!!! What is the difference then? Ask Thomas?}

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (up to $1000$ or less).  UMM 
\ekhine{}{has shown to perform statistically better or equal than CEGO, the state of the art algorithm, in the same instances as in the original paper of CEGO and some new instances. UMM}
is specially well-suited for budget-limited scenarios that are still
time-sensitive, e.g., when the overhead incurred by the optimizer should be no
more than a few minutes added to each fitness evaluation. As shown by our
results, a Bayesian optimizer such as CEGO may sometimes require more than hour
\MANUEL{UMM is able to model the LOP landscape at least as well as CEGO as shown by the plots where UMM shows a linear decrease in fitness.}
\EKHINE{me parece mas de CEGO q UMM y parece q UMM solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos  y luego que addemás es más rápido}
\MANUEL{Pero UMM es peor en rec13 y nug12, y en LOP es mejor porque CEGO no llega a 400, es decir, porque UMM es más rápido. No podemos saber qué pasaría si CEGO llegara a 400. Podríamos medir lo que pasa en el momento en que matamos a CEGO y dar también ese valor entonces la diferencia en velocidad no importaría. }
per fitness evaluation. The time required to train the surrogate-model of CEGO
also sharply increases with permutation length, hence, UMM would be a
computationally feasible alternative for relatively large problem sizes.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to match (and sometimes even improve) the results of a Bayesian
optimizer.
% MANUEL: OK?
% \EKHINE{hhmmm lo hacemos mejor o igual}
The results presented here
show that, despite the intrinsic difficulty of this scenario, there are still
significant advances yet to be made. Thus, we hope that this paper will
motivate further research. % on such problems.

A somewhat surprising result was that the behavior of UMM (and the same for
CEGO) on the PFSP is close to random. This result may be due to the distance
metric used, but further analysis is needed to corroborate this
conjecture. Here, we have focused on Kendall's $\tau$ distance, however, it is
possible to extend UMM to other distance metrics, which will allow us to
dynamically select among various distance metrics for an unknown black-box
% \ekhine{}{estoy totalmente en contra de esto :) }
permutation landscape~\citep{ZaeStoBar2014:ppsn}. Also, as shown by our
experiments, the behavior of the parameters ($r_1$ and $r_2$) of UMM is not
always obvious, thus a more detailed analysis would be needed to provide either
generally good static values or an online adaptation approach. Finally, we have
used here three very different combinatorial problems (LOP, PFSP and QAP) as
black-box benchmarks. However, an even more diverse range of problems would be
needed to understand the behavior of UMM on real-world black-box combinatorial
landscapes.

@@

%% MANUEL: Blind, so no acknowledgments yet.
\begin{smaller}
\paragraph*{Acknowledgements.}
% M.\@ L\'opez-Ib\'a\~nez is a ``Beatriz Galindo'' Senior Distinguished Researcher (BEAGAL 18/00053) funded by the Ministry of Science and Innovation of the Spanish Government.
%
Thanks to Hao Wang (Leiden University) for pointing us to the arguments of
\citet{EriPeaGar2019scalable}.
\end{smaller}

\renewcommand{\doi}[1]{doi:\hspace{.16667em plus .08333em}\discretionary{}{}{}\href{https://doi.org/#1}{\urlstyle{rm}\nolinkurl{#1}}}
\bibliographystyle{splncs04nat}
\bibliography{optbib/abbrev,optbib/authors,optbib/journals,optbib/biblio,optbib/crossref,./mendeley}

>>>>

\newcommand{\supplement}{\href{http://doi.org/10.5281/zenodo.4167148}{doi:10.5281/zenodo.4167148}}

@@

\subsection{Distributions over permutations}
Distributions over permutations~\cite{CriFliVer1996rankings} are functions that assign a probability value  $\Prob(\sigma)\in[0,1]$ to each permutation $\sigma \in S_n$. One of the most popular distributions is the Mallows Model (MM), which is considered as an analogous to the Gaussian distribution for permutations. The MM defines the probability of each permutation $\sigma\in S_n$ as follows:
%
\begin{equation}\label{eq:MM}
\Prob(\sigma \mid \sigma_0, \theta )= \frac{\exp(-\theta d(\sigma, \sigma_0))}{\psi} \sim MM(\sigma_0, \theta)
\end{equation}
%
with two parameters, $\theta$ and $\sigma_0$. Permutation $\sigma_0\in S_n$ is a reference permutation and has the largest probability value, i.e., the mode of the distribution. The probability of every permutation $\sigma\in S_n$ decays exponentially as its distance $d(\sigma,\sigma_0)$ increases, and the dispersion parameter $\theta$  controls this decay.  The normalization constant $\psi$ can be easily computed for the Kendall's $\tau$ distance (Eq.~\eqref{eq:kendall_decomp}) as well as for the Hamming, Cayley and Ulam distances~\cite{IruCalLoz2016permallows}.

@@

The exact MLE for $\sigma_0$ is given by the well-known Kemeny ranking~\cite{DwoKumNao2001rank}. Unfortunately,  obtaining such ranking is NP-hard~\cite{DwoKumNao2001rank}. An alternative to the Kemeny ranking is the Borda ranking~\cite{AliMei2011kemeny}, which has several advantages: (i) it requires polynomial computational time; (ii) it  guaranties  high quality parameters for a sample distributed according to the MM~\cite{CarProSha2013votes}; and (iii) when no distribution is assumed, Borda is an approximation to Kemeny~\cite{CopFleRur2010ordering}.

@@

\subsubsection{Datasets}
%
For the LOP, we generated synthetic instances as follows. First, we create a sample $S$ of 200 permutations of size $n=20$ from a MM with $\sigma_0$ equal to the identity permutation and $\theta=-\log(\textit{Uniformity})$, where  $\textit{Uniformity}\in\{0.5,$ $0.7,$ $0.9\}$, i.e., from peaked distributions to almost uniform distributions. Then, we generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbf{I}\{\sigma(i)<\sigma(j)\}$, where $\mathbf{I}\colon S \times S \to \{0,1\}$ is the indicator function. In principle, these should be best-case instances for UMM since their fitness landscape closely matches the Mallows model.%\MANUEL{Ekhine: Could you revise this text? I wanted to explain why we use synthetic datasets.}

@@

The first part of the experimental analysis concerns the {analyses} of the learning rate, which corresponds to parameter $\rho$ of uBorda. As stated in Section~\ref{sec:sampling}, $\rho$ is automatically set at each iteration in such a way that the $100r_1\%$ of the samples have the $100r_2\%$ of the weight. In this regard, we tried a few different values of parameters $r_1$ and $r_2$ in all the instances. Only values of $r_1 < r_2$ make sense and $r_1 = r_2 = 0.5$ would produce equal weights in the learning step. Thus, we tried values $r_1 = \{0.1, 0.2, \dotsc, 0.5\}$ and $r_2 = \{0.6,\dotsc, 0.9, 0.99\}$. Figure~\ref{fig:heatmaps} shows the best solution found by each independent run (averaged over 10 runs) for instances: \texttt{N-t1d100.01} and \texttt{N-t1d150.01} for the LOP, \texttt{reC05} for the PFSP, \texttt{kra32} for the QAP. We terminated runs that took longer than 6 hours and those combinations are shown as white.  Results for all the datasets can be found in the supplemental material~(\supplement).

@@

The situation is very similar for the QAP, where both algorithms show the same
behaviour as in the PFSP for all instances tested, except for
\texttt{kra32}. As shown in Fig~\ref{fig:kra32}, for \texttt{kra32} both
algorithms are able to improve over the number of evaluations, although not
with ease. As in the synthetic LOP instances, we again observe that CEGO does
improve faster than UMM initially, yet CEGO appears to get stuck around
evaluation 250, and UMM overtakes it around evaluation 300. We believe the
reason is the same as before, that is, the CEGO model is not able to adequately
reflect the ruggedness of the actual landscape and, hence, the best solutions
found by the GA searching the model are not improved solutions for the actual
problem. Interestingly, there are well-known differences between
\texttt{kra32}, on one hand, and \texttt{nug12}, \texttt{nug30} and
\texttt{tho32}, on the other~\citep{AnsBriGou2002qap}. However, we cannot at
this moment point out the root cause of this different behavior.

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (here, up to $400$ evaluations).
UMM \ekhine{}{has shown to perform statistically better or equal than CEGO, the
  state of the art algorithm, in the same instances as in the original paper of
  CEGO and some new instances. UMM} is specially well-suited for budget-limited
scenarios that are still time-sensitive, e.g., when the overhead incurred by
the optimizer should be no more than a few minutes added to each fitness
evaluation. As shown by our results, a Bayesian optimizer such as CEGO may
sometimes require more than hour \MANUEL{UMM is able to model the LOP landscape
  at least as well as CEGO as shown by the plots where UMM shows a linear
  decrease in fitness.}  \EKHINE{me parece mas de CEGO q UMM y parece q UMM
  solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q
  es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos y
  luego que addemás es más rápido} \MANUEL{Pero UMM es peor en rec13 y nug12, y
  en LOP es mejor porque CEGO no llega a 400, es decir, porque UMM es más
  rápido. No podemos saber qué pasaría si CEGO llegara a 400. Podríamos medir
  lo que pasa en el momento en que matamos a CEGO y dar también ese valor
  entonces la diferencia en velocidad no importaría. } per fitness
evaluation. The time required to train the surrogate-model of CEGO also sharply
increases with permutation length, hence, UMM would be a computationally
feasible alternative for relatively large problem sizes.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to match (and sometimes even improve) the results of a Bayesian
optimizer.
% MANUEL: OK?
% \EKHINE{hhmmm lo hacemos mejor o igual}
The results presented here show that, despite the intrinsic difficulty of the
expensive black-box combinatorial scenario, there are still significant
advances yet to be made. Thus, we hope that this paper will motivate further
research. % on such problems.
\MANUEL{I'm thinking about finishing here and commenting out the following paragraph}
A somewhat surprising result was that the behavior of UMM (and the same for
CEGO) on the PFSP is close to random. This result may be due to the distance
metric used, but further analysis is needed to corroborate this
conjecture. Here, we have focused on Kendall's $\tau$ distance, however, it is
possible to extend UMM to other distance metrics, which will allow us to
dynamically select among various distance metrics for an unknown black-box
% \ekhine{}{estoy totalmente en contra de esto :) }
permutation landscape~\citep{ZaeStoBar2014:ppsn}. Also, as shown by our
experiments, the behavior of the parameters ($r_1$ and $r_2$) of UMM is not
always obvious, thus a more detailed analysis would be needed to provide either
generally good static values or an online adaptation approach. Finally, we have
used here three very different combinatorial problems (LOP, PFSP and QAP) as
black-box benchmarks. However, an even more diverse range of problems would be
needed to understand the behavior of UMM on real-world black-box combinatorial
landscapes.

@@

%% MANUEL: Blind, so no acknowledgments yet.
\begin{smaller}
  \paragraph*{Reproducibility.} Source code, datasets and scripts necessary to reproduce the results are available at \supplement.
  \vspace*{-1em}
\paragraph*{Acknowledgements.}
% M.\@ L\'opez-Ib\'a\~nez is a ``Beatriz Galindo'' Senior Distinguished Researcher (BEAGAL 18/00053) funded by the Ministry of Science and Innovation of the Spanish Government.
%
% MANUEL: We should also thank the COST ACTION: Ask Carlos?
Thanks to Hao Wang (Leiden University) for pointing us to the arguments of
\citet{EriPeaGar2019scalable} and to Thomas Stützle (Université libre de Bruxelles) for suggesting works
discussing differences between QAPLIB instances.
\end{smaller}

\renewcommand{\doi}[1]{doi:\hspace{.16667em plus .08333em}\discretionary{}{}{}\href{https://doi.org/#1}{\urlstyle{rm}\nolinkurl{#1}}}
\bibliographystyle{splncs04nat}
\bibliography{optbib/abbrev,optbib/authors,optbib/journals,optbib/biblio,optbib/crossref}%,./mendeley}
